\documentclass[12pt,Bold,letterpaper,TexShade]{mcgilletdclass}

\usepackage{geometry}

\usepackage{texshade}
\usepackage{textcomp}
\usepackage{xspace}
\usepackage{siunitx}
\usepackage{epigraph}
\usepackage{todonotes}
\usepackage{physics}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[labelfont=bf]{caption}
\usepackage{amsmath}
\usepackage{afterpage}






\graphicspath{{figures/}}
%\graphicspath{{figures_lowres/}} 

%hires
%\DeclareGraphicsExtensions{.pdf,.png}
%lores
\DeclareGraphicsExtensions{.png,.pdf}

%%%% VARIABLES %%%%%
\newcommand{\diffusion}{$D_{t}$\xspace}
\newcommand{\gct}{$\gamma$-CT\xspace}
\newcommand{\dcunits}{\cm\squared\per\second}
\newcommand{\figref}[1]{{\bf Fig.~\ref{#1}}}
\newcommand{\vek}[1]{\mathbf{#1}}
\newcommand{\vivo}{{\it in vivo}\xspace}
\newcommand{\silico}{{\it in silico}\xspace}
\newcommand{\vitro}{{\it in vitro}\xspace}
\newcommand{\tub}{$\gamma$-Tubulin\xspace}

%%% SETTINGS %%%
\interfootnotelinepenalty=10000

%\usepackage{tex4ht}
%\usepackage{amsmath}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Have you configured your TeX system for proper  %%
%% page alignment? See the McGillETD documentation %%
%% for two methods that can be used to control     %%
%% page alignment. One method is demonstrated      %%
%% below. See documentation and the ufalign.tex    %%
%% file for instructions on how to adjust these    %%
%% parameters.                                     %%
\addtolength{\hoffset}{0pt}                        %%
\addtolength{\voffset}{0pt}                        %%
%%                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       Define student-specific info
\SetTitle{\huge{Molecular Dynamics of the disordered $\gamma$-tubulin carboxyl terminus}}%
\SetAuthor{Carlos G. Oliver}%
\SetDegreeType{Master of Science}%
\SetDepartment{Department of Biology}%
\SetUniversity{McGill University}%
\SetUniversityAddr{Montreal,Quebec}%
\SetThesisDate{\today}%
\SetRequirements{A thesis submitted to McGill University in partial fulfillment of the requirements of the degree of Master of Science}%
\SetCopyright{\textcopyright  Carlos G. Oliver, 2016}%

\makeindex[keylist]
\makeindex[abbr]


%% Input any special commands below
%\newcommand{\Kron}[1]{\ensuremath{\delta_{K}\left(#1\right)}}
\listfiles%
\begin{document}
\maketitle%

\begin{romanPagenumber}{2}%


\SetDedicationName{\MakeUppercase{Dedication}}%
\SetDedicationText{To my grandparents; Hilda Fernandez and Porfirio Oliver. {\it Gracias por su apoyo, amor, y por ser mi inspiraci\'on en todo.  Los quiero mucho.}}%

\Dedication%

\SetAcknowledgeName{\MakeUppercase{Acknowledgements}}%
\SetAcknowledgeText{I give my sincere thanks to my supervisor Dr. Jackie Vogel for her support and mentorship throughout my Master's, and for sharing with me her passion for understanding the complexity of life. I thank my colleagues in the Vogel lab for their friendship, guidance and useful discussions. I would also like to acknowledge the generous financial support from the Cellular Dynamics of Mollecular Complexes fellowship and Dr. Vogel's grants from the	Canadian Institutes for	Health Research and the National Institutes of Health. I thank the Guillimin support team for their invaluable technical help in working with the Calcul Qu\'ebec compute clusters. I would like to thank Roman Sarrazin Gendron for editing the French abstract. I also thank Vladimir Reinharz for his kind help with computational problems, and for all the enriching discussions.
	\newline \indent  I would like to express my deepest gratitude to all those whom I am so fortunate to call my friends and loved ones. To my brother, whose friendship means everything. To my father for his care and support. And to my mother; for being there for me every day, in every way; for being my role model, and my best friend.}%
\Acknowledge%

\SetContributionName{\MakeUppercase{Contribution of Authors}}%
\SetContributionText{All data collection and results obtained during my M.Sc. are presented here as a standard format thesis. This work includes results and figures obtained by collaborators in the Department of Chemistry Dr. Anthony Mittermaier and Jason Harris who conducted NMR experiments and generated all NMR based figures. All computer simulation work and MD figures were done by myself. Collection of tubulin primary sequences was done by Roman Sarrazin Gendron. All text and literature review in this thesis was done by myself with feedback from my supervisor Jackie Vogel and collaborator Anthony Mittermaier. }%
\Contribution%


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%         English Abstract                        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\SetAbstractEnName{\MakeUppercase{Abstract}}%
\SetAbstractEnText{With recent advances in experimental and computational methods in structural biology, it is becoming increasingly clear that protein function is not only dependent on stable architectures, but equally so on the absence of well-ordered domains. These elements, also known as intrinsically disordered proteins (IDPs) or regions (IDRs) are protein chains that do not adopt organized three dimensional structures, but are highly functional nonetheless. Due to their flexible backbones, IDPs/IDRs are able to explore a large ensemble of conformations and functions. The cell harnesses this flexibility by targeting IDRs for post-translational modifications (PTMs) such as phosphorylation. Altering the conformational sampling of IDRs through PTMs, serves as a useful  tool for functional control and signal integration. This paradigm is crucial for ensuring the proper execution of complex cellular processes that involve a large number of multi-tasking proteins that need to act in a coordinated manner. However, the physical mechanisms linking IDRs to functional output remain largely unknown. In this work, we study the role of phosphorylation in the assembly of the mitotic spindle, which effects chromosome segregation during cell division. Phosphorylation at Tyrosine 11 (Y11) in the intrinsically disordered C-terminus of \tub (\gct) has been shown to control key aspects of the assembly of microtubules in the mitotic spindle. {\it in vivo}, the phosphomimicking mutation of Tyr to Asp (Y11D) leads to a temperature sensitive growth phenotype and an overbuilt spindle in budding yeast. Through Molecular Dynamics computer simulations (MD) and Nuclear Magnetic Resonance spectroscopy (NMR) we show that non-phosphomimicking and phospho-mimicking states of the \gct primarily sample disordered and collapsed conformations. However, the phosphomimicking (Y11D) mutant undergoes switch-like collective motions to an extended but also disordered state. We propose that this transition serves to control the binding of proteins involved in shaping microtubule dynamics. This is the first observation of switch-like behaviour in IDRs/IDPs where both states are disordered, making this a novel physical mechanism of control with the potential to regulate cellular function.}

\AbstractEn%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%         French Abstract                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\SetAbstractFrName{\MakeUppercase{ABR\'{E}G\'{E}}}%
\SetAbstractFrText{De r\'ecents progr\`es dans les m\'ethodologies exp\'erimentales et informatiques en biologie structurelle d\'emontrent que la fonction des prot\'eines ne d\'epend pas uniquement de leur architecture tridimensionnelle stable, mais aussi de l’absence d’une telle architecture. Les \'el\'ements d\'epourvus de structure, appel\'es prot\'eines ou r\'egions intrins\`equement d\'esordonn\'ees (PID, RID), peuvent explorer une vaste gamme de conformations et de fonctions. La cellule exploite cette flexibilit\'e en ciblant les RID avec des modifications post-traductionnelles (MPT) comme la phosphorylation. L’alt\'eration de l'\'echantillonage conformationnel des RID par les MPT est un outil pour le contr\^ole fonctionnel et l’int\'egration des signaux cellulaires. Ce paradigme est essentiel pour assurer l’ex\'ecution correcte des processus cellulaires qui impliquent un grand nombre de prot\'eines multitâches devant agir de mani\`ere pr\'ecise et coordonn\'ee. Cependant, les m\'ecanismes physiques qui \'etablissent un lien entre les RID et les r\'esultats fonctionnels sont en grande partie inconnus.  Cette th\`ese pr\'esente une \'etude sur le r\^ole de la phosphorylation dans l’assemblage et l’organisation du fuseau mitotique, la machine mol\'eculaire qui effectue la s\'egr\'egation des chromosomes durant la division cellulaire. La phosphorylation de l’acide amin\'e Tyrosine 11 (Y11) \`a son extr\'emit\'e carboxy-terminale (\gct) intrins\`equement d\'esordonn\'ee contr\^ole des aspects essentiels de l’assemblage du fuseau mitotique. La mutation de cette tyrosine \`a un acide aspartique phosphomim\'etique \vivo (Y11D) provoque une sensibilit\'e \`a la temp\'erature et des d\'efauts dans l’organisation des microtubules du fuseau mitotique. En utilisant des simulations informatiques de dynamique mol\'eculaire et la r\'esonance magn\'etique nucl\'eaire, nous montrons que les formes non-phosphomim\'etiques et phosphomim\'etiques du \gct balayent toutes deux principalement des conformations compactes et d\'esordonn\'ees, mais que la mutation Y11D entreprend des mouvements collectifs occasionnels vers une conformation mineure \'etendue, elle aussi d\'esordonn\'ee. Nous sugg\'erons que cette transition peut servir \`a contr\^oler la liaison avec les prot\'eines qui affectent la dynamique des microtubules. Il s’agit de la premi\`ere observation d’une transition organis\'ee dans un ensemble de structures d\'esordonn\'ees, un nouveau m\'ecanisme physique de contr\^ole ayant le potentiel de r\'eguler le fonctionnement cellulaire.}%

\AbstractFr%

\TOCHeading{\MakeUppercase{Table of Contents}}%
\LOFHeading{\MakeUppercase{List of Figures}}%
\tableofcontents %
\listoffigures %

\index[abbr]{MD@MD: Molecular Dynamics}
\index[abbr]{\gct@\gct: $\gamma$-Tubulin C-Terminus}
\index[abbr]{NMR@NMR: Nuclear Magnetic Resonance}
\index[abbr]{$\gamma$-TuRC@$\gamma$-TuRC: $\gamma$-Tubulin Ring Complex} 
\index[abbr]{\diffusion@\diffusion: Translational Diffusion Coefficient}
\index[abbr]{WT@WT: Wild-Type}
\index[abbr]{YD@YD: Y445D (Y11D)mutant}
\index[abbr]{RMSD@RMSD: Root Mean Square Displacement}
\index[abbr]{$R_g$@$R_g$: Radius of Gyration}
\index[abbr]{ns@ns: Nanosecond}
\index[abbr]{$\mu$s@$\mu$s: Microsecond}
\index[abbr]{IDP@IDP: Intrinsically Disordered Protein}
\index[abbr]{IDR@IDR: Intrinsically Disordered Region}
\index[abbr]{CREB@CREB: Cyclic AMP Response Element Binding Protein}
\index[abbr]{CBP@CBP: CREB Binding Protein}
\index[abbr]{KID@KID: Kinase Inducible Domain}
\index[abbr]{DNA@DNA: Deoxyribonucleic Acid}
\index[abbr]{GRIPS@GRIPS: $\gamma$-Tubulin Ring Proteins}
\index[abbr]{PTM@PTM: Post-Translational Modification}
\index[abbr]{NVE@NVE: Microcanonical Ensemble}
\index[abbr]{NVT@NVT: Canonical Ensemble}
\index[abbr]{NPT@NPT: Isothermal-isobaric Ensemble}
\index[abbr]{OPLS-AA@OPLS-AA: Optimized Potential for Liquid Simulations - All Atom}
\index[abbr]{SPCE@SPCE: Extended Single Point Charge}
\index[abbr]{GROMACS@GROMACS: GROnigen Machine for Chemical Simulatoins}
\index[abbr]{VMD@VMD: Visual Molecular Dynamics}


\printindex[abbr]{KEY TO ABBREVIATIONS}{KEY TO ABBREVIATIONS}{}

\end{romanPagenumber}

%\mainmatter %

%introduction chapter
\input{introduction.tex} 

%methods chapter
\input{methods.tex}

\input{gctmd_chapter.tex}

\chapter{Conclusions}
\input{conclusion.tex}

\input{appendix.tex}
\bibHeading{References}
\bibliography{thesis_cgo}
\bibliographystyle{plain}




\end{document}


 






